
Because Milkys purports to be Norway's best source for pollution in the marine environment, it merits some closer inspection. In this file I will work through both data from and the reports behind Milkys, to develop a better understanding of the dataset and campaigns.

I will use as sources here the following documents and data:

1. A Vannmiljø search (all pollutants, campaign == Milkys, no other filtering)
2. NIVA's accompanying Contaminants in coastal waters reports, 2020-2024 (inclusive, so 5 reports)

```{r}
library(arrow)
library(tidyverse)
library(here)
library(sf)
library(knitr)

vm_milkys_all <- read_parquet(
  "inst/example_datasets/registrations/WaterRegistrationExport-Milkys-AllPollutants.parquet"
)

```


```{r}
vm_milkys_all |>
  group_by(year = year(Tid_provetak)) |>
  reframe(
    n = n(),
    n_stressors = n_distinct(Parameter_id),
    n_sites = n_distinct(Vannlokalitet_kode),
    media = str_flatten_comma(sort(unique(Medium_navn)))
  ) |>
  arrange(desc(year))
```

So, obvious point here: almost no non-biota media were sampled. In fact, Sediment saltvann is present for only one year in the whole period. This matches what we (and AI) see in the report: 

> The 2024 optional monitoring also included sediment cores (Oslo, Bergen, and Tromsø) 

So at least as far as concerns sediment and seawater, Milkys is no use to us. Which is a shame, but it's nice to get an unequivocal no. 