---
title: "Milkys"
description: >
  A very quick review of Milkys - it only contains biota data.
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
---

Because we assumed Milkys to be the best source for pollution in the marine environment, it merits some closer inspection. In this file I will work through both data from and the reports behind Milkys, to develop a better understanding of the dataset and campaigns.

I will use as sources here the following documents and data:

1. A Vannmiljø search (all pollutants, campaign == Milkys, no other filtering)
2. NIVA's accompanying Contaminants in coastal waters reports, 2020-2024 (inclusive, so 5 reports)

```{r}
#| echo: false
library(arrow)
library(tidyverse)
library(here)
library(knitr)

vm_milkys_all <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "registrations",
    "WaterRegistrationExport-Milkys-AllPollutants.parquet"
  )
)
```


```{r}
vm_milkys_all |>
  group_by(year = year(Tid_provetak)) |>
  reframe(
    n = n(),
    n_stressors = n_distinct(Parameter_id),
    n_sites = n_distinct(Vannlokalitet_kode),
    media = str_flatten_comma(sort(unique(Medium_navn)))
  ) |>
  arrange(desc(year)) |>
  kable()
```

So, obvious point here: almost no non-biota media were sampled. In fact, Sediment saltvann is present for only one year in the whole period. This matches what we (and AI) see in the report: 

> The 2024 optional monitoring also included sediment cores (Oslo, Bergen, and Tromsø) 

So at least as far as concerns sediment and seawater, Milkys is no use to us. Which is a shame, but it's nice to get an unequivocal no. 