Vannmiljø's Aktivitet names/IDs correspond both to formal monitoring campaigns with associated reports (e.g. MILKYS), as well as other sources of data. Vannmiljø gives a list of activities in its [kodeverk website](https://vannmiljokoder.miljodirektoratet.no/activity?q=) (undated). 

In their own words:

> All miljøovervåking og -kartlegging har som utgangspunkt å fremskaffe tilstrekkelig kunnskap til å kunne utøve en best mulig forvaltning av naturmiljøet. Det enkelte overvåkingsprosjekt eller -program er innrettet for å oppnå mer spesifikke forvaltningsmål, f.eks. å kontrollere at avbøtende tiltak faktisk fører til forbedring i miljøtilstanden. For å sikre en best mulig utnyttelse av ressursene til overvåking av miljøtilstanden i vann, er det viktig å ha oversikt over alle pågående og avsluttede overvåkingsaktiviteter. Det gir bedre grunnlag for å samordne aktiviteter gjennom felles prøvetaking og sammenholde ferske data med resultater fra tidligere overvåking. Derfor skal alle overvåkingsdata (vannregistreringer) i Vannmiljø knyttes til en overvåkingsaktivitet. En vannregistrering kan bare tilhøre en overvåkingsaktivitet, mens en vannlokalitet kan ha mange vannregistreringer med forskjellige aktiviteter.

> I tabellen nedenfor er det gitt en oversikt over alle gyldige aktivitetskoder (ActivityID) med utfyllende beskrivelse av aktivitetene. Listen er ment å skulle være så uttømmende som mulig, men kan utvides ved behov. Vi vil imidlertid være restriktive med stadige utvidelser siden det motvirker formålet med listen, nemlig å gi oversikt.

Let's do a quick exploration of the available list of activities. I'm using a local version from 2026 here, although you could also get a live version from the API, if you wanted.

```{r}
library(arrow)
library(tidyverse)
library(here)
library(sf)
library(knitr)

vm_activities <- read_parquet(
  "inst/example_datasets/codelists/Vannmiljo_Aktivitet_2026-10-05.parquet"
)

vm_activities |> kable()

```

These descriptions are a useful starting point. Some important points:

1. Not all activities are of equal importance
2. Some activities focus on the same (or similar) sites each year (e.g. reference activities such as `Basisovervåking - påvirka områder`/`Basisovervåking - referanseforhold`, which are designed to provide long-term data on reference contaminated and uncontaminated sites)
3. Some activities are formal sampling campaigns conducted on behalf of the NEA (e.g. [Milkys](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/forurensning-og-klimagasser/miljogifter-langs-kysten/)), and have yearly reports and lots of metadata
4. Other activities are data collected by other sources for other purposes, and are stored in Vannmiljø because it's the easiest way to make such data available (even if it isn't relevant to water)

So, we come into this with a bunch of assumptions that we need to test. In order to properly understand the data, we need to review both the datasets we have and available reporting about monitoring campaigns and other activities. This is easier said than done.

Because the scope of the work I'm currently doing is associated with Arctic/marine ecosystems, we will focus on the campaigns that most contribute to these datasets. We've defined our criteria elsewhere:

1. Medium = `Saltvann` or `Sediment saltvann`
2. Latitude `< 66.5636` (once reprojected to CRS 4326)

Also as elsewhere, we'll examine the data structure from several subsets.

## Sediment Saltvann, 2010 - 2025

`WaterRegistrationExport_NO_Jan10_Jan25_SV` represents the data extracted from Vannmiljø's `Søk i miljøgifter` frontend for Jan 01 2010 - Jan 01 2025, with `Medium` set to `Saltvann`. It is assumed that this data is a representative sample. 

The returned file has 83,277 rows. Once we filter it to Arctic sites we're working with about 10%: 7,168 rows.

```{r}
library(arrow)

# 83,277 rows
WaterRegistrationExport_NO_Jan10_Jan25_SV <- read_parquet(
  "inst/example_datasets/registrations/WaterRegistrationExport-NO-Jan10-Jan25-SV.parquet"
)

reproject_4326 <- function(dataset) {
  dataset |>
    filter(
      !is.na(`UTM33 Ost (X)`) &
        !is.na(`UTM33 Nord (Y)`)
    ) |>
    st_as_sf(
      coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
      crs = 25833,
      remove = FALSE
    ) |>
    st_transform(4326) |>
    mutate(LATITUDE = st_coordinates(geometry)[, 2])
}

WaterRegistrationExport_NO_Jan10_Jan25_SV_reproj <- WaterRegistrationExport_NO_Jan10_Jan25_SV |>
  reproject_4326() |>
  mutate(arctic = LATITUDE >= arctic_circle_lat)
# drop rows without spatial data, 82,467 rows

WaterRegistrationExport_NO_Jan10_Jan25_SV_reproj |>
  filter(arctic) |>
  st_drop_geometry() |>
  reframe(.by = Aktivitet_navn, n = n(), Aktivitet_id = unique(Aktivitet_id)) |>
  arrange(desc(n))
```

Here we get some bad news. The single biggest dataset is `Mikronor` / `MILP`, which monitors microplastics. We're only interested 



# Basisovervåking

According to https://www.vannportalen.no/kunnskapsgrunnlaget/overvaking2/basisovervaking/,
Basisovervåking consists of the following activities:

Kystvann:  

    Økosystemovervåking i kystvann (Økokyst) (miljødirektoratet)
    Miljøgifter i kystområdene (Milkys) (miljødirektoratet) ()
    Havforsuringsprogrammet (miljødirektoratet)

Ferskvann (elver og innsjøer):

    Overvåking av elvetilførsler og direkte utslipp til norske kystområder (Elvetilførselsprogrammet): 
        Elveovervåkningsprogrammet – vannkvalitetsstatus og trender
        Kildefordelte tilførsler av nitrogen og fosfor til norske kystområder
    Økosystemovervåking i store sjøer (ØKOSTOR)
    Økosystemovervåking i ferskvann (ØKOFERSK)
    Referanseelver
    Grunnvann


# Aquaculture

Aquaculture facilities are required to conduct frequent monitoring of the sediment below
their sites for contamination. In some areas (such as the Arctic), this is the single largest source of pollution data, and eclipses formal monitoring campaigns. 

https://www.vannportalen.no/veiledere/veiledning-i-bruk-av-data-fra-miljoundersokelser-ved-akvakulturanlegg-i-sjo-til-klassifisering-av-vannforekomster/



However, when we've looked at actual Vm data (see for example the horizontal/vertical slices), basisovervåkning is its own activity.

## Milkys

Milkys is the most obvious source of pollution data relevant to the Arctic/marine, but does that assumption hold?

Using our copper data from the other exploration, let's visualise.

```{r copper-load-data}
# slightly different format to the more recent extracted data above!

library(arrow)
library(tidyverse)
library(here)
library(sf)

copper_registrations <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "registrations",
    "Vm_Copper_2025.12.05.parquet"
  )
)

copper_sites <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "sites",
    "Vm_Copper_Sites_2025.12.05.parquet"
  )
)

copper_dataset <- left_join(
  copper_registrations,
  copper_sites,
  by = join_by(Vannlok_kode == Vannlokalitetskode)
)

campaign_lookup <- read_csv(here(
  "data/raw/vannmiljo/Vm_lookup_campaigns_SAW_2024.csv"
)) |>
  select(CAMPAIGN, CAMPAIGN_DESCRIPTION_NO) |>
  rename(Aktivitet_id = CAMPAIGN, Aktivitet_navn = CAMPAIGN_DESCRIPTION_NO)

copper_dataset_reproj <- copper_dataset |>
  filter(
    !is.na(`UTM33 Ost (X)`) &
      !is.na(`UTM33 Nord (Y)`)
  ) |>
  st_as_sf(
    coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
    crs = 25833,
    remove = FALSE
  ) |>
  st_transform(4326) |>
  mutate(LATITUDE = st_coordinates(geometry)[, 2])

arctic_circle_lat <- 66.5636 # approximate latitude of the Arctic Circle

highlight_colours <- c(
  "Other" = "#c3c2b7",
  "Arctic" = "#2a78d6",
  "Marine (Kyst)" = "#eb6834",
  "Arctic & marine" = "#1baf7a"
)

copper_dataset_highlights <- copper_dataset_reproj |>
  mutate(
    arctic = LATITUDE >= arctic_circle_lat,
    # this time we don't have a type field...
    marine = ((Medium_id %in% c("VS", "SS", "PSS", "NS")) | # saltvann, Sediment saltvann, Porevann sediment saltvann, Bunnsubstrat saltvann
      (str_detect(Vannlok_kode, "HAV"))),
    highlight = case_when(
      arctic & marine ~ "Arctic & marine",
      arctic ~ "Arctic",
      marine ~ "Marine (Kyst)",
      .default = "Other"
    ),
    highlight = factor(highlight, levels = names(highlight_colours))
  ) |>
  left_join(campaign_lookup, by = join_by(Aktivitet_id))
```

Come to think of it, which campaigns look at which compartments? (copper again):

```{r}
copper_dataset |>
  group_by(Aktivitet_id, Medium_id) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  left_join(campaign_lookup, by = "Aktivitet_id") |>
  mutate(
    is_relevant_medium = Medium_id %in% c("SV", "SS", "VS", "PSS"),
    is_interesting_campaign = Aktivitet_id %in%
      c("CEMP", "MILK", "TILT", "AKVA"),
    Medium_id <- fct_inorder(Medium_id),
    Aktivitet_navn = fct_inorder(Aktivitet_navn)
  ) |>

  ggplot(aes(y = Aktivitet_navn, x = Medium_id, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Log sample size") +
  facet_grid(
    cols = vars(is_relevant_medium),
    ,
    rows = vars(is_interesting_campaign),
    scales = "free",
    space = "free"
  )
```

This time let's just look at Arctic data

```{r copper-campaign-representation}
#| fig-cap: "Copper monitoring data points per campaign, 2010-2025."

copper_campaign_rep <- copper_dataset_highlights |>
  group_by(Aktivitet_id, Aktivitet_navn, highlight) |>
  filter(highlight == "Arctic & marine") |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  mutate(Aktivitet_navn = fct_inorder(Aktivitet_navn) |> fct_rev())


ggplot(
  data = copper_campaign_rep,
  aes(y = Aktivitet_navn, fill = highlight, x = n)
) +
  geom_col() +
  scale_fill_manual(values = highlight_colours, name = NULL) +
  labs(title = "Copper monitoring data points by campaign; 2010-2025")
```

So yeah, really not a lot of Milkys. Actuallt, according to the below table, there's only 7 data points! 

```{r}
copper_campaign_rep
```

If we count both Milkys and CEMP (Milkys' old name) that still only gives 167 data points that are Artic/marine for the period. Which is suspicious. So let's try an alternative.

Specifically salt water (`Medium == Saltvann`)


Milkys isn't even on here! I guess it doesn't do water. 