---
title: "Vanmniljø Arctic/Marine Exploration"
description: >
  Exploring data relevant to the EXPECT case study, starting with Tromsø.
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    lightbox: true
    fig-dpi: 300
knitr:
  opts_chunk:
    dev: ragg_png
execute:
  warning: false
  message: false
---

Through a process of not exactly scientific prioritisation (see vannmiljø-expect-data.qmd) we've arrived at a more-or-less reasonable list of areas which I think merit futher attention. I'm desperate to start working with more specific locations, because I think doing stuff at the national level has a lot of challenges. We'll work with the same arctic-saltwater/saltwater sediment dataset we have before:


```{r}
#| warning: false
library(arrow)
library(here)
library(dplyr)
library(sf)
library(ggplot2)
library(readxl)
library(ggrepel)
library(patchwork)
```

```{r}
WaterRegistrationExport_NO_Jan10_Jan25_SV <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "registrations",
    "WaterRegistrationExport-NO-Jan10-Jan25-SV.parquet"
  )
)

WaterRegistrationExport_NO_Jan10_Jan25_SedSV <- read_parquet(
  here(
    "inst/example_datasets/registrations/WaterRegistrationExport-NO-Jan10-Jan25-SedSV.parquet"
  )
)

pollutants <- read_excel(
  here("data/raw/vannmiljo/Vannmiljø_Miljøgifter_2026-09-29.xlsx")
)

all_data <- add_row(
  WaterRegistrationExport_NO_Jan10_Jan25_SV,
  WaterRegistrationExport_NO_Jan10_Jan25_SedSV
)

arctic_circle_lat <- 66.5636 # approximate latitude of the Arctic Circle

reproject_4326 <- function(dataset) {
  dataset |>
    filter(
      !is.na(`UTM33 Ost (X)`) &
        !is.na(`UTM33 Nord (Y)`)
    ) |>
    st_as_sf(
      coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
      crs = 25833,
      remove = FALSE
    ) |>
    st_transform(4326) |>
    mutate(LATITUDE = st_coordinates(geometry)[, 2])
}

all_data_reproj <- all_data |>
  reproject_4326() |>
  mutate(arctic = LATITUDE >= arctic_circle_lat)
# drop rows without spatial data

all_data_reproj_arctic <- all_data_reproj |> filter(arctic)

```

In my analysis of the expect-specific data I identified the following sites:

Counting place names in the prose (code chunks excluded), these are the ones the document mentions most:

| Place | Mentions |
|---|---|
| Tromsø | 7 |
| Hammerfest | 6 |
| Harstad | 5 |
| Narvik | 4 |
| Alta | 3 |
| Bodø | 3 |
| Vadsø | 3 |
| Finnsnes | 2 |
| Kaldsletta | 2 |
| Tysfjorden | 2 |
| Fauske, Kirkenes, Lenvikk, Mosjøen, Svolvær | 1 each |

It seems the pragmatic approach would be to start with Tromsø and see how long it takes.

```{r}
map_crs <- 25833 # ETRS89 / UTM 33N, in metres
bigger_arctic_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 65, xmax = 40, ymax = 75),
  crs = sf::st_crs(4326)
)

norway_cities <- read_excel(here("inst", "worldcities.xlsx")) |>
  filter(iso2 == "NO") |>
  st_as_sf(coords = c("lng", "lat"), crs = st_crs(4326)) |>
  st_crop(bigger_arctic_bbox) |>
  filter(st_coordinates(geometry)[, 2] >= arctic_circle_lat - 1) |>
  st_transform(map_crs)
```

# Tromsø

I'm going to use a totally arbitrary starting approach of drawing a 60 x 60km box around Tromsø and seeing what it hits. Panels show the water framework directive (WFD) water bodies, then the number of samples and number of distinct pollutants in 3 km grid squares, for seawater and for sediment. A sample here is a unique combination of site, sampling time, depth and sample number.

Let's set some bounding boxes:
- 60km
- 20km 
- 10km

We don't immediately know which of these may be the most appropriate.

```{r}
tromso_centre <- norway_cities |>
  filter(city == "Tromsø") |>
  st_geometry() |>
  st_coordinates()

# half_width is in km; map_crs units are metres
st_point_bbox <- function(point, half_width, map_crs) {
  half_width <- half_width * 1000
  st_bbox(
    c(
      xmin = point[1] - half_width,
      xmax = point[1] + half_width,
      ymin = point[2] - half_width,
      ymax = point[2] + half_width
    ),
    crs = st_crs(map_crs)
  )
}


tromso_box_60 <- st_point_bbox(tromso_centre, 30, 25833)
tromso_box_20 <- st_point_bbox(tromso_centre, 10, 25833)
tromso_box_10 <- st_point_bbox(tromso_centre, 5, 25833)

tromso_box <- st_as_sfc(tromso_box_60) # the 60 km box used for the grid below

# the smaller boxes as sf polygons, so they can be overlaid on maps
tromso_boxes <- st_sf(
  box = c("20 km", "10 km"),
  geometry = c(st_as_sfc(tromso_box_20), st_as_sfc(tromso_box_10))
)
```

```{r}
wfd_bodies <- read_sf(here(
  "inst",
  "shapefiles",
  "vannforekomster_kyst_arctic.gpkg"
)) |>
  st_transform(map_crs) |>
  st_crop(tromso_box_60) |>
  arrange(navn) |>
  mutate(wfd_no = sprintf("%02d", row_number()))

land <- rnaturalearth::ne_countries(
  scale = "large",
  continent = "Europe",
  returnclass = "sf"
) |>
  st_transform(map_crs) |>
  st_crop(tromso_box_60)

tromso_cities <- norway_cities |> st_filter(tromso_box)

# 3 km grid over the box (cell numbers are per cell, not per site)
grid <- st_make_grid(tromso_box, cellsize = 3000) |>
  st_sf(geometry = _) |>
  mutate(cell = row_number())

tromso_data <- all_data_reproj |>
  st_transform(map_crs) |>
  st_filter(tromso_box)

grid_summary <- tromso_data |>
  st_join(grid) |>
  st_drop_geometry() |>
  filter(!is.na(cell)) |>
  reframe(
    .by = c(cell, Medium_navn),
    samples = n_distinct(Vannlokalitet_kode, Tid_provetak, Ovre_dyp, Provenr),
    pollutants = n_distinct(Parameter_navn)
  )

grid_long <- grid_summary |>
  tidyr::pivot_longer(
    c(samples, pollutants),
    names_to = "metric",
    values_to = "n"
  ) |>
  mutate(
    metric = factor(
      metric,
      c("samples", "pollutants"),
      c(
        "Number of samples (3km grid square)",
        "Number of pollutants (3km grid square)"
      )
    )
  ) |>
  left_join(grid, by = "cell") |>
  st_as_sf()

base_map <- function() {
  list(
    geom_sf(data = land, fill = "grey92", colour = "grey60", linewidth = 0.2),
    geom_sf(data = tromso_cities, shape = 21, colour = "red", size = 1.5),
    coord_sf(
      xlim = tromso_box_60[c("xmin", "xmax")],
      ylim = tromso_box_60[c("ymin", "ymax")],
      expand = FALSE
    ),
    theme_minimal(base_size = 9),
    theme(axis.text = element_blank(), panel.grid = element_blank())
  )
}
city_labels <- geom_text_repel(
  data = tromso_cities,
  aes(label = city, geometry = geometry),
  stat = "sf_coordinates",
  size = 2.5,
  bg.colour = "white",
  bg.r = 0.15,
  max.overlaps = Inf
)


density_panel <- function(metric, medium, option) {
  ggplot() +
    geom_sf(
      data = wfd_bodies,
      fill = "#d0ffff",
      colour = "grey40",
      linewidth = 0.15
    ) +
    geom_sf(
      data = grid_long |> filter(metric == !!metric, Medium_navn == medium),
      aes(fill = n),
      colour = NA,
      alpha = 0.75
    ) +
    geom_sf(
      data = tromso_boxes,
      aes(colour = box),
      fill = NA,
      linewidth = 0.8
    ) +
    scale_fill_viridis_c(option = option, name = NULL) +
    labs(title = paste(metric, "-", medium))
}
```

## WFD water bodies

There are 43 WFD water bodies in our bounding box. This is probably a sign that we've drawn it too big. It would make the most sense (I think) to zoom to the middle 

```{r}
#| fig-width: 6
#| fig-height: 6

p_wfd <- ggplot() +
  geom_sf(
    data = wfd_bodies,
    aes(fill = kjemiskTilstand),
    alpha = 0.7,
    colour = "grey30",
    linewidth = 0.2
  ) +
  geom_sf_text(
    data = suppressWarnings(st_centroid(wfd_bodies)),
    aes(label = wfd_no),
    size = 2.5,
    fontface = "bold"
  ) +
  geom_sf(
    data = tromso_boxes,
    aes(colour = box),
    fill = NA,
    linewidth = 0.8
  ) +
  scale_colour_manual(values = c("20 km" = "red", "10 km" = "blue")) +
  city_labels +
  labs(
    title = "WFD water bodies, 60km bounding box centered Tromsø",
    fill = "Chemical status",
    colour = "Box"
  )

p_wfd
```

```{r}
wfd_bodies |>
  st_drop_geometry() |>
  select(wfd_no, navn, vannforekomst_id, kjemiskTilstand)
```

## Saltvann

```{r}
#| fig-width: 12
#| fig-height: 5
density_panel("Number of samples (3km grid square)", "Saltvann", "viridis") +
  density_panel("Number of pollutants (3km grid square)", "Saltvann", "magma")
```

## Sediment saltvann

```{r}
#| fig-width: 12
#| fig-height: 5
density_panel(
  "Number of samples (3km grid square)",
  "Sediment saltvann",
  "viridis"
) +
  density_panel(
    "Number of pollutants (3km grid square)",
    "Sediment saltvann",
    "magma"
  )
```
