---
title: "Vanmniljø Arctic/Marine Exploration"
description: >
  Exploring data relevant to the EXPECT case study.
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    lightbox: true
    dpi: 300
execute:
  warning: false
  message: false
---

Based on the scoping and exploration work we've done elsewhere we can start to focus on our main area of interest: mixtures of pollutants above the Arctic circle, in saltwater and saltwater sediment. We'll use the datasets that are filtered to `Saltvann` and `Saltvann sediment` simply because that's the easiest way to get a relevant subset of the data from Vannmiljø. 

```{r}
#| warning: false
library(arrow)
library(here)
library(dplyr)
library(sf)
library(ggplot2)
library(readxl)
library(ggrepel)
```

```{r}
WaterRegistrationExport_NO_Jan10_Jan25_SV <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "registrations",
    "WaterRegistrationExport-NO-Jan10-Jan25-SV.parquet"
  )
)

WaterRegistrationExport_NO_Jan10_Jan25_SedSV <- read_parquet(
  here(
    "inst/example_datasets/registrations/WaterRegistrationExport-NO-Jan10-Jan25-SedSV.parquet"
  )
)

pollutants <- read_excel(
  here("data/raw/vannmiljo/Vannmiljø_Miljøgifter_2026-09-29.xlsx")
)

all_data <- add_row(
  WaterRegistrationExport_NO_Jan10_Jan25_SV,
  WaterRegistrationExport_NO_Jan10_Jan25_SedSV
)

nrow(all_data)
```

Now filter to above the Arctic circle:


```{r}
arctic_circle_lat <- 66.5636 # approximate latitude of the Arctic Circle

reproject_4326 <- function(dataset) {
  dataset |>
    filter(
      !is.na(`UTM33 Ost (X)`) &
        !is.na(`UTM33 Nord (Y)`)
    ) |>
    st_as_sf(
      coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
      crs = 25833,
      remove = FALSE
    ) |>
    st_transform(4326) |>
    mutate(LATITUDE = st_coordinates(geometry)[, 2])
}

all_data_reproj <- all_data |>
  reproject_4326() |>
  mutate(arctic = LATITUDE >= arctic_circle_lat)
# drop rows without spatial data

all_data_reproj_arctic <- all_data_reproj |> filter(arctic)

nrow(all_data_reproj_arctic)
```

Roughly 20% of the data is reportedly Arctic. Let's visualise and check:

```{r}
all_data_reproj_sf <- all_data_reproj |>
  group_by(Vannlokalitet_kode) |>
  reframe(
    Type,
    Medium_navn,
    Aktivitet_navn,
    Parameter_navn,
    arctic,
    geometry
  ) |>
  st_as_sf()

norway_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 57.5, xmax = 35.5, ymax = 81),
  crs = sf::st_crs(4326)
)

europe_context <- rnaturalearth::ne_countries(
  scale = "large",
  continent = "Europe",
  returnclass = "sf"
) |>
  st_crop(
    c(xmin = 4, ymin = 57.5, xmax = 45, ymax = 81) |>
      st_bbox() |>
      st_set_crs(4326)
  )

ggplot(europe_context) +
  geom_sf() +
  geom_hline(
    yintercept = arctic_circle_lat,
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_sf(data = all_data_reproj_sf, aes(colour = arctic))
```

That looks right. Let's look at the hotspots where stressors are measured the most:


```{r}
bigger_arctic_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 65, xmax = 40, ymax = 75),
  crs = sf::st_crs(4326)
)

arctic_context <- europe_context

# Hexes are binned in x/y units, so project to metres first (ETRS89 / UTM 33N)
map_crs <- 25833
plot_bbox <- bigger_arctic_bbox |>
  st_as_sfc() |>
  st_transform(map_crs) |>
  st_bbox()

# Trim the empty east: a lon/lat box widens once projected, so set xmax from
# the projected extent of the arctic data (plus a 50 km margin) instead.
plot_bbox[["xmax"]] <- all_data_reproj_sf |>
  filter(arctic) |>
  st_transform(map_crs) |>
  st_coordinates() |>
  _[, 1] |>
  max() +
  50000

# A parallel is a curve in UTM 33N, so build it as a densified line in lon/lat
# and project it, rather than using a straight horizontal line.
arctic_circle_line <- st_linestring(
  cbind(seq(0, 45, by = 0.25), arctic_circle_lat)
) |>
  st_sfc(crs = 4326) |>
  st_transform(map_crs)

# Only label the larger cities that are on the map and near/north of the Arctic
# Circle; the rest are mostly noise relative to the hexes.
norway_cities <- read_excel(here("inst", "worldcities.xlsx")) |>
  filter(iso2 == "NO") |>
  st_as_sf(coords = c("lng", "lat"), crs = st_crs(4326)) |>
  st_crop(bigger_arctic_bbox) |>
  filter(st_coordinates(geometry)[, 2] >= arctic_circle_lat - 1) |>
  st_transform(map_crs)

all_data_reproj_sf_xy <- all_data_reproj_sf |>
  st_crop(bigger_arctic_bbox) |>
  st_transform(map_crs) |>
  mutate(
    x = st_coordinates(geometry)[, 1],
    y = st_coordinates(geometry)[, 2]
  )

# Where the projected Arctic Circle meets the right-hand edge, for its label
arctic_label_y <- st_coordinates(arctic_circle_line) |>
  as.data.frame() |>
  slice_min(abs(X - plot_bbox[["xmax"]]), n = 1) |>
  pull(Y)

plot_density <- function(data, medium_navn, option, title) {
  ggplot() +
    geom_sf(
      data = arctic_context,
      fill = NA,
      colour = "grey30",
      linewidth = 0.2
    ) +
    geom_hex(
      data = if (is.null(medium_navn)) {
        data
      } else {
        filter(data, Medium_navn == medium_navn)
      },
      aes(x = x, y = y),
      bins = 60,
      alpha = 0.75
    ) +
    geom_sf(
      data = arctic_circle_line,
      colour = "firebrick",
      linetype = "dashed",
      linewidth = 0.6
    ) +
    annotate(
      "text",
      x = plot_bbox[["xmax"]],
      y = arctic_label_y,
      label = "Arctic Circle",
      colour = "firebrick",
      hjust = 1,
      vjust = -0.5,
      size = 3.2
    ) +
    geom_sf(
      data = norway_cities,
      shape = 21,
      fill = NULL,
      colour = "red",
      size = 1.8
    ) +
    geom_text_repel(
      data = norway_cities,
      aes(label = city, geometry = geometry),
      stat = "sf_coordinates",
      size = 3,
      colour = "black",
      bg.colour = "white", # white halo keeps labels legible over the hexes
      bg.r = 0.15,
      min.segment.length = 0.2,
      segment.colour = "grey40",
      max.overlaps = Inf
    ) +
    coord_sf(
      crs = map_crs,
      xlim = plot_bbox[c("xmin", "xmax")],
      ylim = plot_bbox[c("ymin", "ymax")],
      expand = FALSE
    ) +
    scale_fill_viridis_c(
      name = "Measurements",
      option = option
    ) +
    theme_minimal() +
    theme(axis.title = element_blank()) +
    labs(title = title)
}
```

```{r}
plot_density(
  all_data_reproj_sf_xy,
  "Saltvann",
  "viridis",
  "Monitoring density, arctic sites: saltvann"
)
```

What do we learn? Well, first of all, we don't have a lot of choices when it comes to saltwater monitoring sites. Vadsø is by far the most monitored site, followed by:

- Hammerfest
- Harstad
- Tromsø/Kaldsletta
- Tysfjorden, SW of Narvik

```{r}
plot_density(
  all_data_reproj_sf_xy,
  "Sediment saltvann",
  "plasma",
  "Monitoring density, arctic sites: sediment saltvann"
)
```

## Without aquaculture?

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn != "Miljøovervåking akvakulturanlegg"),
  "Sediment saltvann",
  "plasma",
  "Monitoring density, arctic sites: sediment saltvann"
)
```

Sediment monitoring is much more thorough across the entire coast of Norway, and at points in the sea. The most monitored sites are:

- Tromsø/Kaldsletta
- Harstad
- Bodø
- Hammerfest

This raises some immediate questions:

1. Why is Vadsø important for sea water but not sediment?
2. Likewise Tysfjorden


What are the main campaigns contributing here again?


```{r}
# this is still sediment and seawater only

all_data_reproj |>
  filter(arctic) |>
  st_drop_geometry() |>
  reframe(
    .by = "Aktivitet_navn",
    n = n(),
    n_stressors = n_distinct(Parameter_casnr)
  ) |>
  arrange(desc(n))
```

## Density by activity

One section per activity, ordered by number of measurements (n, from the table above).

### Overvåking av forurenset sjøbunn

n = 49602.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av forurenset sjøbunn"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av forurenset sjøbunn"
)
```

### Kartlegging av miljøgifter i sedimenter - MAREANO

n = 40808.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(
      Aktivitet_navn == "Kartlegging av miljøgifter i sedimenter - MAREANO"
    ),
  "Sediment saltvann",
  "viridis",
  "Kartlegging av miljøgifter i sedimenter - MAREANO"
)
```

### Miljøovervåking akvakulturanlegg

n = 12273.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Miljøovervåking akvakulturanlegg"),
  "Sediment saltvann",
  "viridis",
  "Miljøovervåking akvakulturanlegg"
)
```

### Overvåking av påvirkning fra industri

n = 5899.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av påvirkning fra industri"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av påvirkning fra industri"
)
```

### Annet

n = 5770.

```{r}
plot_density(
  all_data_reproj_sf_xy |> filter(Aktivitet_navn == "Annet"),
  "Sediment saltvann",
  "viridis",
  "Annet"
)
```

### Tiltaksorientert overvåking

n = 3199.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Tiltaksorientert overvåking"),
  "Sediment saltvann",
  "viridis",
  "Tiltaksorientert overvåking"
)
```

### Mikroplast i kystområder, elver og innsjøer (Mikronor)

n = 3098.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(
      Aktivitet_navn == "Mikroplast i kystområder, elver og innsjøer (Mikronor)"
    ),
  "Sediment saltvann",
  "viridis",
  "Mikroplast i kystområder, elver og innsjøer (Mikronor)"
)
```

### Effekter av mudring, utfylling og dumping

n = 2936.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Effekter av mudring, utfylling og dumping"),
  "Sediment saltvann",
  "viridis",
  "Effekter av mudring, utfylling og dumping"
)
```

### Overvåking av avrenning fra landdeponi

n = 971.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av avrenning fra landdeponi"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av avrenning fra landdeponi"
)
```

### Miljøgifter i kystområdene (MilKys)

n = 938.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Miljøgifter i kystområdene (MilKys)"),
  "Sediment saltvann",
  "viridis",
  "Miljøgifter i kystområdene (MilKys)"
)
```

### Effekter av planlagt arealbruk

n = 919.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Effekter av planlagt arealbruk"),
  "Sediment saltvann",
  "viridis",
  "Effekter av planlagt arealbruk"
)
```

### Problemkartlegging

n = 896.

```{r}
plot_density(
  all_data_reproj_sf_xy |> filter(Aktivitet_navn == "Problemkartlegging"),
  "Sediment saltvann",
  "viridis",
  "Problemkartlegging"
)
```

### Overvåking av påvirkning fra flyplasser

n = 693.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av påvirkning fra flyplasser"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av påvirkning fra flyplasser"
)
```

### Overvåking av påvirkning fra avløp

n = 583.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av påvirkning fra avløp"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av påvirkning fra avløp"
)
```

### Tilførselsprogrammet

n = 578.

```{r}
plot_density(
  all_data_reproj_sf_xy |> filter(Aktivitet_navn == "Tilførselsprogrammet"),
  "Sediment saltvann",
  "viridis",
  "Tilførselsprogrammet"
)
```

### Overvåking av påvirkning fra vegtrafikk

n = 465.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Overvåking av påvirkning fra vegtrafikk"),
  "Sediment saltvann",
  "viridis",
  "Overvåking av påvirkning fra vegtrafikk"
)
```

### Myndighetspålagt forurensningsovervåking

n = 445.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Myndighetspålagt forurensningsovervåking"),
  "Sediment saltvann",
  "viridis",
  "Myndighetspålagt forurensningsovervåking"
)
```

### Basisovervåking - påvirka områder

n = 324.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Basisovervåking - påvirka områder"),
  "Sediment saltvann",
  "viridis",
  "Basisovervåking - påvirka områder"
)
```

### Feltspesifikk miljøovervåking på norsk sokkel

n = 308.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Feltspesifikk miljøovervåking på norsk sokkel"),
  "Sediment saltvann",
  "viridis",
  "Feltspesifikk miljøovervåking på norsk sokkel"
)
```

### Kartlegging av nye miljøgifter

n = 148.

```{r}
plot_density(
  all_data_reproj_sf_xy |>
    filter(Aktivitet_navn == "Kartlegging av nye miljøgifter"),
  "Sediment saltvann",
  "viridis",
  "Kartlegging av nye miljøgifter"
)
```

### Watch List

n = 147.

```{r}
plot_density(
  all_data_reproj_sf_xy |> filter(Aktivitet_navn == "Watch List"),
  "Sediment saltvann",
  "viridis",
  "Watch List"
)
```


```{r}
cell_size <- 20000 # metres (UTM 33N)

plot_grid_density <- function(data, title, option = "viridis") {
  ggplot() +
    geom_sf(
      data = arctic_context,
      fill = NA,
      colour = "grey30",
      linewidth = 0.2
    ) +
    geom_bin_2d(
      data = data,
      aes(x = x, y = y),
      binwidth = c(cell_size, cell_size),
      boundary = 0, # same grid origin on every map, so maps are comparable
      alpha = 0.75
    ) +
    geom_sf(
      data = arctic_circle_line,
      colour = "firebrick",
      linetype = "dashed",
      linewidth = 0.6
    ) +
    geom_sf(
      data = norway_cities,
      shape = 21,
      fill = "white",
      colour = "black",
      size = 1.8
    ) +
    geom_text_repel(
      data = norway_cities,
      aes(label = city, geometry = geometry),
      stat = "sf_coordinates",
      size = 3,
      bg.colour = "white",
      bg.r = 0.15,
      min.segment.length = 0.2,
      segment.colour = "grey40",
      max.overlaps = Inf
    ) +
    coord_sf(
      crs = map_crs,
      xlim = plot_bbox[c("xmin", "xmax")],
      ylim = plot_bbox[c("ymin", "ymax")],
      expand = FALSE
    ) +
    scale_fill_viridis_c(name = "Measurements", option = option) +
    theme_minimal() +
    theme(axis.title = element_blank()) +
    labs(title = title)
}

plot_grid_density(
  all_data_reproj_sf_xy |>
    filter(
      Medium_navn == "Sediment saltvann",
      Aktivitet_navn == "Overvåking av forurenset sjøbunn"
    ),
  "Overvåking av forurenset sjøbunn: 20 km grid"
)

```


```{r}
# Arctic coastal water bodies (see data-raw/prepare_vannforekomster.R)
water_bodies <- read_sf(
  here("inst", "shapefiles", "vannforekomster_kyst_arctic.gpkg")
) |>
  st_transform(map_crs) |>
  select(navn, vannforekomst_id)

# Collapse the names of the water bodies a cell overlaps into one string,
# e.g. "Vestfjorden, Skjerstadfjorden (+2 more)"
summarise_overlaps <- function(idx, max_names = 3) {
  nms <- unique(water_bodies$navn[idx])
  if (length(nms) == 0) {
    return("none")
  }
  shown <- paste(head(nms, max_names), collapse = ", ")
  if (length(nms) > max_names) {
    shown <- paste0(shown, " (+", length(nms) - max_names, " more)")
  }
  shown
}

cell_table <- function(data, cell_size = 10000, top_n = 15) {
  cells <- data |>
    st_drop_geometry() |>
    mutate(
      cell_x = floor(x / cell_size) * cell_size,
      cell_y = floor(y / cell_size) * cell_size
    ) |>
    summarise(
      measurements = n(),
      stressors = n_distinct(Parameter_navn),
      sites = n_distinct(Vannlokalitet_kode),
      .by = c(cell_x, cell_y)
    ) |>
    slice_max(measurements, n = top_n)

  # cell centres (for nearest city) and square cell polygons (for water bodies)
  centres <- st_as_sf(
    mutate(cells, cx = cell_x + cell_size / 2, cy = cell_y + cell_size / 2),
    coords = c("cx", "cy"),
    crs = map_crs
  )
  cell_polygons <- st_buffer(
    centres,
    dist = cell_size / 2,
    endCapStyle = "SQUARE"
  )

  nearest <- st_nearest_feature(centres, norway_cities)
  overlaps <- st_intersects(cell_polygons, water_bodies)

  cells |>
    mutate(
      nearest_city = norway_cities$city[nearest],
      km_to_city = round(
        as.numeric(
          st_distance(centres, norway_cities[nearest, ], by_element = TRUE)
        ) /
          1000
      ),
      water_bodies = vapply(overlaps, summarise_overlaps, character(1)),
      per_site = round(measurements / sites, 1),
      share = scales::percent(measurements / nrow(data), accuracy = 0.1)
    ) |>
    select(
      nearest_city,
      km_to_city,
      water_bodies,
      measurements,
      sites,
      per_site,
      share
    )
}

all_data_reproj_sf_xy |>
  filter(
    Medium_navn == "Sediment saltvann",
    Aktivitet_navn == "Overvåking av forurenset sjøbunn"
  ) |>
  cell_table() |>
  knitr::kable()
```
