---
title: "Vannmiljø Data Exploration - Arctic & Marine"
description: >
  Exploring subsets of Vannmiljø data to see how much information they could give us about mixtures of pollutants in Arctic marine environments.
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-overflow: wrap
    lightbox: true
error: true
---

```{r setup}
#| include: false

library(tidyverse)
library(readxl)
library(here)
library(arrow)
library(sf)
library(rnaturalearth)
library(purrr)
library(scales)
library(leaflet)
library(knitr)
library(ggrepel)
```

Vannmiljø data is divided into various monitoring campaigns ("Aktivitet"), multi-year programs of monitoring specific ecosystems, pressures, etc. These are listed here: https://vannmiljokoder.miljodirektoratet.no/activity?q=.

These campaigns are relatively well-described, but it's not always clear what they cover. I've gone through and assigned them a likely relevance score based on descriptions, but let's take a look at the actual data too.

# Definitions

Our brief for this task is:

1. Mixtures of pollutants
2. Arctic sites
3. Marine sites

Let's clarify these:

## Mixtures of pollutants

Per last meeting, we are specifically interested in traditional ecotox chemical pollutants (e.g. heavy metals, organic chemistry, halogen chemistry, organometallic chemistry), but not anything ecological, quality parameters, nutrients, etc. We will use Vannmiljø's own `Miljøgifter` category to make this decision.

## Arctic Sites

Anything above 66.5636 N. This is a minority of Norway, both by population and area.

## Marine sites

This is slightly more complicated. Vannmiljø stores this data in a variety of formats, and they aren't always consistent. I will clarify marine sites to "salt/transitional water", and sediment/porewater from the same. In Vannmiljø, this information is stored as follows:

`Vannlokalitet_kode` associated with a site may start `HAV-`, which indicates that the site is an ocean.

`Type` may be marked `Kyst`. This indicates that a site is coastal, and is (I think) mutually exclusive with `Vannlokalitet_kode` being a `HAV`

`Medium_navn` and/or `Medium_id` may be one of the saltwater media. I have defined these as follows, from the [Vm medium codelist](https://vannmiljokoder.miljodirektoratet.no/medium?q=salt):

| Code | Name | Description |
|------|------|-------------|
| NS | Bunnsubstrat saltvann | Marin bløtbunnsfauna og marin hardbunnsflora og -fauna. |
| PSS | Porevann sediment saltvann | Vannvolum av porevann fra vertikale snitt i sedimentert materiale i marint vann. |
| SS | Sediment saltvann | Vertikale snitt av sedimentert materiale i marint vann. Analyseresultater angis i tørrvekt. |
| VS | Saltvann | Vannvolum fra overflate eller nærmere angitte dyp/dybdeintervall i marint vann. Omfatter også marine organismer i pelagialen og på bløtbunn og hardbunn. |

I haven't fully investigated the crossover between these categories. I hope that together, they are sufficiently discriminatory to highlight relevant groups.

Important detail: I match these using `OR` rather than `AND`. So we may capture e.g. biota samples (which would have a different `Type`) to those listed above, because they're sampled from a `HAV-...` site.

# Horizontal Slice

Let's start with a horizontal slice of Vannmiljø, hopefully more-or-less representative. This is all the data from 01 January 2020 to 31 May 2020, which is 464,059 rows.

```{r load-data}
horizontal_slice <- read_parquet(here(
  "inst",
  "example_datasets",
  "registrations",
  "WaterRegistrationExport-NO-Jan20-May20.parquet"
))

pollutants <- read_excel(
  here("data", "raw", "vannmiljo", "Vannmiljø_Miljøgifter_2026-09-29.xlsx")
)

parameters <- unique(pollutants$Name)
```

## Environmental Pollutants in Horizontal Slice

What sort of pollutants are in Vannmiljø?

`QualityElementName` only tells us if a pollutant is `Prioriterte stoffer` or `Andre stoffer`. This might be useful.

```{r pollutant-quality-element}
pollutants |>
  group_by(QualityElementName) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  kable()
```

`SubGroupName` is more useful, and splits substances as follows:

```{r pollutant-subgroups}
pollutants |>
  group_by(SubGroupName, SubGroupID) |>
  reframe(n = n(), example = first(Name)) |>
  arrange(desc(n)) |>
  kable()
```

This is a lot of groups, and each group has a lot of members. There's no obvious group to pick on as the most relevant. 

How well represented are these pollutants in our slice? (note that this step excludes anything that Vannmiljø doesn't consider a pollutant.)

```{r pollutant-representation-join}
pollutant_representation <- right_join(
  horizontal_slice,
  pollutants,
  by = join_by(Parameter_id == ParameterID)
)
```

- First, most important note. Right-joining to pollutants leaves us with 75,320 rows, less than a quarter of the original data.



Which subgroups are best represented?


::: {.callout-note}

## A quick digression - highlighting relevant sites


`UTM33 Ost (X)`/`UTM33 Nord (Y)` are projected coordinates (ETRS89 / UTM zone 33N, EPSG:25833), so we can't threshold them against a latitude directly — we need to reproject to WGS84 first (same as `edata_sites()` does for the site export, see `R/fct_vm_eData.R`).

```{r reproject-coordinates}
reprojected <- pollutant_representation |>
  filter(
    !is.na(pollutant_representation$`UTM33 Ost (X)`) &
      !is.na(pollutant_representation$`UTM33 Nord (Y)`)
  ) |>
  st_as_sf(
    coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
    crs = 25833,
    remove = FALSE
  ) |>
  st_transform(4326) |>
  mutate(LATITUDE = st_coordinates(geometry)[, 2])
```

Now flag the arctic and marine rows once, on `pollutant_representation` itself, so every plot downstream groups by the same `highlight` categories instead of recomputing them.

How do we decide if a site is "Marine". The following criteria:

```{r arctic-marine-flags}
arctic_circle_lat <- 66.5636 # approximate latitude of the Arctic Circle

highlight_colours <- c(
  "Other" = "#c3c2b7",
  "Arctic" = "#2a78d6",
  "Marine (Kyst)" = "#eb6834",
  "Arctic & marine" = "#1baf7a"
)

pollutant_representation <- reprojected |>
  mutate(
    arctic = LATITUDE >= arctic_circle_lat,
    marine = Type == "Kyst" |
      str_detect(Medium_navn, "[sS]alt") |
      str_detect(Vannlokalitet_kode, "HAV"),
    highlight = case_when(
      arctic & marine ~ "Arctic & marine",
      arctic ~ "Arctic",
      marine ~ "Marine (Kyst)",
      .default = "Other"
    ),
    highlight = factor(highlight, levels = names(highlight_colours))
  )
```

Do our classifications make sense when we map them out?

```{r classification-map}
#| fig-cap: "Sites classified as Arctic, marine, both, or neither, mapped against the Arctic Circle (dashed line)."

site_sf <- pollutant_representation |>
  group_by(Vannlokalitet_kode) |>
  reframe(Type, highlight, geometry) |>
  st_as_sf()

norway_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 57.5, xmax = 35.5, ymax = 81),
  crs = sf::st_crs(4326)
)

europe_context <- rnaturalearth::ne_countries(
  scale = "large",
  continent = "Europe",
  returnclass = "sf"
) |>
  filter(sovereignt != "Russia") |>
  st_crop(norway_bbox)

ggplot(europe_context) +
  geom_sf() +
  geom_hline(
    yintercept = arctic_circle_lat,
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_sf(data = site_sf, aes(colour = highlight)) +
  scale_colour_manual(values = highlight_colours)
```

Broadly speaking, yes: marine sites are in the ocean, and arctic sites are above the arctic circle. 

Now that we have these classifications, we can start looking at what sort of information do we have for these sites.

:::

```{r subgroup-representation}
#| fig-cap: "Number of samples per pollutant subgroup, coloured by Arctic/marine classification."

subgroup_rep <- pollutant_representation |>
  group_by(SubGroupID, SubGroupName, highlight) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  mutate(SubGroupName = fct_inorder(SubGroupName) |> fct_rev())

ggplot(
  data = subgroup_rep,
  mapping = aes(x = n, y = SubGroupName, fill = highlight)
) +
  geom_col() +
  scale_fill_manual(values = highlight_colours, name = NULL) +
  theme_minimal()
```

We can see already that it's not looking great: metals are the most commonly monitored pollutant, but even for them less than 10% of samples are from marine ecosystems. Next up is PAHs, organic chemistr, and PFAS, the proportion of which are marine/arctic is <1%.

It's worth taking a closer look at each class of chemical.

```{r subgroup-parameter-plots}
#| layout-ncol: 2

subgroups <- unique(pollutant_representation$SubGroupName)

walk(subgroups, .f = function(subgroup) {
  plot_data <- pollutant_representation |>
    filter(SubGroupName == subgroup) |>
    group_by(Parameter_id, Parameter_navn, highlight) |>
    reframe(n = n()) |>
    group_by(Parameter_id) |>
    mutate(total_n = sum(n)) |>
    ungroup() |>
    arrange(desc(total_n)) |>
    mutate(
      Parameter_id = fct_inorder(Parameter_id),
      Parameter_id = fct_lump_n(Parameter_id, 30, w = total_n) |> fct_rev()
    )

  p <- ggplot(
    data = plot_data,
    mapping = aes(x = Parameter_id, y = n, fill = highlight)
  ) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = highlight_colours, name = NULL) +
    scale_y_continuous(breaks = breaks_pretty()) +
    labs(
      title = subgroup,
      subtitle = "Most common parameter per group, lumped after 30 groups"
    ) +
    theme_minimal()

  print(p)
})
```


## Arctic + Marine Subset

Now that we've identified the most relevant subset of the data, it's worth looking at what campaigns contribute most to this datast:

```{r campaign-representation}
#| fig-cap: "Number of samples per monitoring campaign, coloured by Arctic/marine classification."

campaign_rep <- pollutant_representation |>
  group_by(Aktivitet_navn, Aktivitet_id, highlight) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  mutate(Aktivitet_navn = fct_inorder(Aktivitet_navn) |> fct_rev())

ggplot(data = campaign_rep, aes(y = Aktivitet_navn, fill = highlight, x = n)) +
  geom_col() +
  scale_fill_manual(values = highlight_colours, name = NULL)
```

Based on my reading of this data, it seems like our most relevant campaigns are:

```{r arctic-marine-relevant-campaigns}
campaign_rep |>
  filter(highlight == "Arctic & marine" & n > 0) |>
  kable()
```

Perhaps surprisingly, aquaculture monitoring is far and away the best single source.

## Sites with mixtures

Another assumption I want to test - how many sites do we have where mixtures are measured? 

Starting with the Arctic/marine subset (of 785 data points!):

```{r multi-stressor-site-summary}
multi_stressors_arctic_marine_sites <- pollutant_representation |>
  filter(highlight == "Arctic & marine") |>
  group_by(Vannlokalitet_kode, Vannlokalitetsnavn) |>
  reframe(
    n = n(),
    n_stressors = n_distinct(Parameter_navn),
    media = str_flatten_comma(unique(Medium_navn)),
    stressors = str_flatten_comma(unique(Parameter_navn)),
    geometry = unique(geometry)
  ) |>
  arrange(desc(n)) |>
  st_as_sf()
```

I've cropped this map aggressively; note that this may not hold for different datasets.

Note: Sites filtered to n_stressors > 10

This is a pretty shoddy map, but I hope it gets the message across: there don't seem to be a lot of good candidate sites in this map.

```{r multi-stressor-arctic-site-maps}
#| fig-cap: "Arctic marine sites where more than 10 distinct stressors were measured, Jan-May 2020."
#| warning: false

multi_stressors_arctic_marine_sites_gt_10 <- multi_stressors_arctic_marine_sites |>
  filter(n_stressors > 10)

arctic_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 67, xmax = 20, ymax = 69.8),
  crs = sf::st_crs(4326)
)

arctic_context <- europe_context |>
  st_crop(arctic_bbox)

ggplot(arctic_context) +
  geom_sf() +
  geom_sf(
    data = multi_stressors_arctic_marine_sites_gt_10,
    aes(colour = media)
  ) +
  geom_label_repel(
    arrow = arrow(length = unit(0.01, "npc")),
    data = multi_stressors_arctic_marine_sites_gt_10,
    aes(label = Vannlokalitet_kode, geometry = geometry),
    stat = "sf_coordinates",
    max.overlaps = Inf,
    min.segment.length = 0,
    force = 10,
    force_pull = 0.1
  ) +
  labs(
    title = "Arctic marine sites where > 10 stressors were measured",
    subtitle = "Jan 01 to May 31 2020"
  )
```

```{r multi-stressor-site-table}
multi_stressors_arctic_marine_sites_gt_10 |>
  st_drop_geometry() |>
  kable()
```

# Vertical Slice

To make my life easier, I'm going to assume that copper, being the most studied stressor in Norway, is a fairly good proxy for the overall monitoring of an area. 

This dataset covers all copper concentrations measured from 2010 - 2025. Note that the formatting is slightly different to the more recently extracted dataset above. I'm not sure if this is due to format changes or using the UI slightly differently; in either case it's worrying.

```{r copper-load-data}
# slightly different format to the more recent extracted data above!

copper_registrations <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "registrations",
    "Vm_Copper_2025.12.05.parquet"
  )
)

copper_sites <- read_parquet(
  here(
    "inst",
    "example_datasets",
    "sites",
    "Vm_Copper_Sites_2025.12.05.parquet"
  )
)

copper_dataset <- left_join(
  copper_registrations,
  copper_sites,
  by = join_by(Vannlok_kode == Vannlokalitetskode)
)

campaign_lookup <- campaign_rep |>
  select(Aktivitet_navn, Aktivitet_id) |>
  distinct_all()

copper_dataset_reproj <- copper_dataset |>
  filter(
    !is.na(`UTM33 Ost (X)`) &
      !is.na(`UTM33 Nord (Y)`)
  ) |>
  st_as_sf(
    coords = c("UTM33 Ost (X)", "UTM33 Nord (Y)"),
    crs = 25833,
    remove = FALSE
  ) |>
  st_transform(4326) |>
  mutate(LATITUDE = st_coordinates(geometry)[, 2])


copper_dataset_highlights <- copper_dataset_reproj |>
  mutate(
    arctic = LATITUDE >= arctic_circle_lat,
    # this time we don't have a type field...
    marine = ((Medium_id %in% c("VS", "SS", "PSS", "NS")) | # saltvann, Sediment saltvann, Porevann sediment saltvann, Bunnsubstrat saltvann
      (str_detect(Vannlok_kode, "HAV"))),
    highlight = case_when(
      arctic & marine ~ "Arctic & marine",
      arctic ~ "Arctic",
      marine ~ "Marine (Kyst)",
      .default = "Other"
    ),
    highlight = factor(highlight, levels = names(highlight_colours))
  ) |>
  left_join(campaign_lookup, by = join_by(Aktivitet_id))
```

And visualise:

```{r copper-campaign-representation}
#| fig-cap: "Copper monitoring data points per campaign, 2010-2025."

copper_campaign_rep <- copper_dataset_highlights |>
  group_by(Aktivitet_id, Aktivitet_navn, highlight) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  mutate(Aktivitet_navn = fct_inorder(Aktivitet_navn) |> fct_rev())

# why is it not in order?
ggplot(
  data = copper_campaign_rep,
  aes(y = Aktivitet_navn, fill = highlight, x = n)
) +
  geom_col() +
  scale_fill_manual(values = highlight_colours, name = NULL) +
  labs(title = "Copper monitoring data points by campaign; 2010-2025")
```

```{r copper-arctic-marine-campaigns}
copper_campaign_rep |>
  filter(highlight == "Arctic & marine" & n > 0) |>
  kable()
```

So most campaigns over the period have *some* arctic/marine data, but it's very much in the minority? `Miljøovervåking akvakulturanlegg`  and `Overvåking av forurenset sjøbunn` are still the most important groups by an order of magnitude.

## Monitoring Hotspots

We obviously can't directly measure which sites have mixtures with just copper, but it's worth looking at where the most frequently monitored locations in the Arctic area: presumably this will hold for mixtures, too.

Coastal sites in Vannmiljø are identified by a location code starting `AA.BB`, which corresponds to a given fjord in the fjord catalog. 

```{r copper-hotspot-hexmap}
#| fig-cap: "Density of copper monitoring records across Arctic/marine sites, 2010-2025."

# copper_dataset_highlights_sf <- copper_dataset_highlights |>
#   filter(highlight == "Arctic & marine") |>
#   group_by(Vannlok_kode, Vannlokalitetsnavn) |>
#   reframe(
#     n = n(),
#     media = str_flatten_comma(unique(Medium_id)),
#     lat = unique(st_coordinates(geometry)[, 2]),
#     lon = unique(st_coordinates(geometry)[, 1])
#   ) |>
#   arrange(desc(n))

copper_dataset_highlights_sf <- copper_dataset_highlights |>
  filter(highlight == "Arctic & marine") |>
  mutate(
    lat = (st_coordinates(geometry)[, 2]),
    lon = (st_coordinates(geometry)[, 1])
  )

bigger_arctic_bbox <- sf::st_bbox(
  c(xmin = 4, ymin = 65, xmax = 40, ymax = 75),
  crs = sf::st_crs(4326)
)

arctic_context <- europe_context |>
  st_crop(bigger_arctic_bbox)

ggplot() +
  geom_hex(
    data = copper_dataset_highlights_sf,
    aes(x = lon, y = lat),
    alpha = 0.9,
    bins = 75
  ) +
  geom_sf(data = arctic_context, fill = NA) +
  coord_sf(crs = 4326) +
  scale_fill_viridis_c(name = "n") +
  labs(title = "Copper monitoring density, arctic sites")
```

`Vannlokalitetsnavn` is, on the whole, not very useful. This potentially becomes a reverse geo-coding issue. 

Hotspots seem to include:

1. Bodø and the fjord to the east
2. Harstad (probably the single biggest concentration of sites)
3. Tromsø
4. Hammerfest

The table below shows only the 30 most common sites... there are 5,797 in total.

```{r copper-site-name-representation}
copper_dataset_highlights_sf |>
  group_by(Vannlok_kode, Vannlokalitetsnavn) |>
  reframe(n = n()) |>
  arrange(desc(n)) |>
  slice_max(n = 30, order_by = n) |>
  kable()
```

I want to get some advice here because I feel this is quite an important question: how do we group sites together in a way that's a) scalable, b) scientific, and c) effective.?

<!-- Actually, a leaflet map might be the shout here, since it already has place data and clusters automatically. (disabled for rendering) -->

```{r copper-leaflet-map}
# pal <- colorFactor(highlight_colours, domain = names(highlight_colours))

# leaflet(copper_dataset_highlights_sf) |>
#   addTiles() |>
#   addCircleMarkers(
#     color = ~ pal(highlight),
#     radius = 5,
#     stroke = FALSE,
#     fillOpacity = 0.8,
#     popup = ~ paste0("<b>", Vannlokalitetsnavn, "</b><br>", Vannlok_kode),
#     clusterOptions = markerClusterOptions()
#   ) |>
#   addLegend(pal = pal, values = ~highlight, title = NULL)
```

### Fjord IDs

One option is to use fjord IDs. These look like `04.21.02.10.00` == `Akkarfjorden`; Vannlok_kode's first two elements for coastal sites correspond to the first two groups of the parent fjord. This may or may not be useful...?

```{r}
#| eval: false
fjord_id <- read_csv("inst/shapefiles/fjord-catalog/fjord_catalog_lookup.csv")
```

If we have only the first two elements to match on:

```{r}
#| eval: false
fjord_id_short <- fjord_id |>
  mutate(id_short = substr(fjordid, start = 1, stop = 5))

fjord_id_short |>
  group_by(id_short) |>
  reframe(n = n(), names = str_flatten_comma(unique((navn)))) |>
  filter(!is.na(names))
```

Ok, I'm making the call that we can't just group by fjord ID. 

### Cities

I don't have a good understanding of cities in Norway, so let's get some data.

```{r}
norway_cities <- read_excel(here("inst", "worldcities.xlsx")) |>
  filter(iso2 == "NO") |>
  mutate(
    bigger_city = population > 20000
  ) |>
  st_as_sf(coords = c("lng", "lat"), crs = st_crs(4326))
```

We know that Bodo is relatively well-characterised, so what if we just look there?

```{r copper-hotspot-bodo}
arctic_context <- europe_context

# Hexes are binned in x/y units, so project to metres first (ETRS89 / UTM 33N)
map_crs <- 25833
plot_bbox <- bigger_arctic_bbox |>
  st_as_sfc() |>
  st_transform(map_crs) |>
  st_bbox()

copper_xy <- copper_dataset_highlights_sf |>
  st_crop(bigger_arctic_bbox) |>
  st_transform(map_crs) |>
  mutate(
    x = st_coordinates(geometry)[, 1],
    y = st_coordinates(geometry)[, 2]
  )

ggplot() +
  geom_hex(
    data = copper_xy,
    aes(x = x, y = y),
    alpha = 0.9,
    bins = 75
  ) +
  geom_sf(data = arctic_context, fill = NA) +
  geom_sf(data = norway_cities |> st_crop(bigger_arctic_bbox), colour = "red") +
  geom_sf_text(
    data = norway_cities |> st_crop(bigger_arctic_bbox),
    colour = "red",
    aes(label = city),
    hjust = 0, # text starts at the point and runs right
    vjust = 1 # text hangs below the point
  ) +
  # crs = map_crs so the non-sf hex layer (already in metres) lines up;
  # limits are then also in metres
  coord_sf(
    crs = map_crs,
    xlim = plot_bbox[c("xmin", "xmax")],
    ylim = plot_bbox[c("ymin", "ymax")]
  ) +
  scale_fill_viridis_c(name = "n") +
  labs(title = "Copper monitoring density, arctic sites")

copper_dataset_highlights_sf |>
  st_crop(bigger_arctic_bbox) |>
  group_by(Vannlok_kode, Aktivitet_id) |>
  reframe(n = n()) |>
  arrange(desc(n))
```

Does arctic + marine hold consistent over time (in the vertical slice), or does it vary? If the latter, we can't assume that the horizontal slice is representative.

Next tasks:

- Look at the last 5 years only.
- Review hotspots - population centres, plus individual fjords with activities
- Get a basic summary of sources, etc. 
- Find areas with the most contamination, overlap with protection
- Ask Claude: which Arctic areas are known to be highly polluted
- But also would be helpful from an arctic perspective to identify the most monitored pollutants in the Arctic
- Characterise mixtures at hotspots
- Only consider sediment and water concentrations
- Send KET a list of relevant reports
- Get 

= Could we do the whole-artic data and have a look at how much spatial location explains it?

= Could scope it to water and sediment, sørfjorden
- Do some AI summaries of relevant monitoring reports/campaigns

1. Get Claude to review NEA reports over the past 5 years for Arctic data
2. Use this to select hotspots (sediment and water, do some sort of gridding exercise\)
3. Review those hotspots in the data - what mixtures do you encounter?

Also: get this document exported properly to HTML/PDF
